---
title: "Feature Blocks, Subspaces, and Multi-Stage Pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature Blocks, Subspaces, and Multi-Stage Pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette covers three advanced capabilities of neuralign:

1. **Feature blocks** --- composing multiple correspondence signals into a
   single alignment input, with harmonization and weighting.
2. **Subspace operations** --- working in the identified subspace when
   correspondence features are rank-deficient.
3. **Multi-stage pipelines** --- composing sequential alignment models into a
   single transform.

```{r setup}
library(neuralign)
```

## Part 1: Feature blocks

### The problem

Alignment methods need a correspondence signal: a set of measurements shared
across subjects that defines "what should match." In practice, you often have
*multiple* signals---task activation betas from different conditions, anatomical
features, connectivity fingerprints---each with different reliability,
coverage, and independence properties.

### Conventions: when transforms act on rows vs. columns

neuralign applies alignment transforms by **left-multiplication**:
`X_aligned = T %*% X`, where `X` is `features × observations`.

Some workflows instead estimate a transform that acts on the **columns**
(`X_aligned = X %*% T`). In neuralign, this distinction is controlled by the
`convention` argument in diagnostics/builders:

- `convention = "left"`: the transform dimension is `nrow(X)` (acts on rows).
- `convention = "right"`: the transform dimension is `ncol(X)` (acts on columns).

If your correspondence matrices are naturally in a right-multiply form (e.g.,
`P × K` matrices where you want a `K × K` basis rotation), you can still use
`fit_alignment()` by transposing them to `K × P`. See
`vignette("fmri_profile", package = "neuralign")` for a worked example.

Feature blocks provide a structured way to:

- represent each signal as a named, weighted block,
- harmonize blocks across subjects (handling missing features),
- stack blocks into a single matrix for alignment.

### Creating blocks

```{r create-blocks}
set.seed(1)

# Task activation betas (3 conditions x 8 observations)
task_block <- alignment_feature_block(
  x = matrix(rnorm(24), 3, 8),
  name = "task_beta",
  weight = 1,
  feature_names = c("face", "scene", "object"),
  meta = list(
    source_type = "supervised",
    requires_independence = TRUE
  )
)

# Anatomical features (2 features x 8 observations)
anat_block <- alignment_feature_block(
  x = matrix(rnorm(16), 2, 8),
  name = "anat",
  weight = 0.5,
  feature_names = c("thickness", "curvature"),
  meta = list(
    source_type = "prior",
    requires_independence = FALSE
  )
)

str(task_block, max.level = 1)
```

**Weight semantics:** `stack_feature_blocks()` scales each block by
`sqrt(weight)`. This means weights behave naturally in sum-of-squares
objectives (Procrustes, etc.). If your legacy pipeline applied a linear
weight `w`, the equivalent neuralign weight is `w^2`.

### Harmonizing across subjects

When subjects have different feature vocabularies (e.g., different brain
regions survive quality control), `harmonize_feature_blocks()` reconciles
them:

```{r harmonize-blocks}
set.seed(2)

blocks_by_subject <- list(
  s1 = list(
    task = alignment_feature_block(
      matrix(rnorm(24), 3, 8), "task",
      feature_names = c("face", "scene", "object")
    ),
    anat = alignment_feature_block(
      matrix(rnorm(16), 2, 8), "anat",
      feature_names = c("thickness", "curvature")
    )
  ),
  s2 = list(
    task = alignment_feature_block(
      matrix(rnorm(24), 3, 8), "task",
      feature_names = c("object", "face", "scene")  # different order
    ),
    anat = alignment_feature_block(
      matrix(rnorm(16), 2, 8), "anat",
      feature_names = c("curvature", "thickness")  # different order
    )
  )
)

harmonized <- harmonize_feature_blocks(blocks_by_subject, min_features = 2)
# Features are now in consistent order across subjects
```

**Intersection mode** (default): keeps only features present in all subjects.
This is the conservative default.

### Stacking blocks

After harmonization, stack blocks into a single matrix per subject:

```{r stack-blocks}
F_s1 <- stack_feature_blocks(harmonized$s1)
F_s2 <- stack_feature_blocks(harmonized$s2)

dim(F_s1)  # (n_task + n_anat) x n_obs, with sqrt(weight) scaling
rownames(F_s1)  # prefixed: "task:face", "task:scene", ..., "anat:thickness", ...
```

### The build_alignment_features() pipeline

For production use, `build_alignment_features()` combines harmonization,
stacking, coverage tracking, and identifiability checking in one call:

```{r build-features}
features <- build_alignment_features(
  blocks_by_subject,
  harmonize = "intersection",
  min_features = 2,
  check_identifiability = TRUE,
  convention = "right"
)

names(features)
features$per_block
features$identifiability
```

### Union-fill mode

When features are sparse across subjects, intersection may drop too many
features. Union-fill mode keeps the full vocabulary and fills missing entries
with zeros:

```{r union-fill}
blocks_sparse <- list(
  s1 = list(
    task = alignment_feature_block(
      matrix(rnorm(24), 3, 8), "task",
      feature_names = c("face", "scene", "object")
    )
  ),
  s2 = list(
    task = alignment_feature_block(
      matrix(rnorm(16), 2, 8), "task",
      feature_names = c("scene", "tool")  # only 'scene' overlaps
    )
  )
)

features_union <- build_alignment_features(
  blocks_sparse,
  harmonize = "union_fill",
  fill = 0,
  warn_sparse_below = 0.5
)

# s2 now has zero-filled rows for "face" and "object"
dim(features_union$matrices$s1)
dim(features_union$matrices$s2)
features_union$coverage
```

### Independence tracking

Blocks that are estimated from the data (e.g., task betas) should be flagged
with `meta$requires_independence = TRUE`. When building features, neuralign
warns if independence is not maintained:

```{r independence-check}
b <- alignment_feature_block(
  matrix(rnorm(12), 3, 4), "task",
  feature_names = c("f1", "f2", "f3"),
  meta = list(requires_independence = TRUE)
)

# This warns because obs_crossfit = FALSE (default)
suppressWarnings(
  build_alignment_features(
    list(s1 = list(task = b), s2 = list(task = b)),
    check_independence = TRUE,
    obs_crossfit = FALSE
  )
)
```

Set `obs_crossfit = TRUE` to suppress the warning when you have properly
cross-fit your correspondence signals.

## Part 2: Identifiability and subspaces

### The identifiability problem

When your correspondence signal has fewer effective dimensions than the
transform space, only part of the transformation is **identified**---the rest
is unconstrained and may overfit.

Example: if you have 5 correspondence features (task conditions) but 100
voxels, only a 5-dimensional subspace of the 100-dimensional transformation
space is constrained by the data.

### Diagnosing identifiability

```{r identifiability}
set.seed(3)

# Small correspondence signal (3 features x 10 observations)
# in a 10-dimensional transform space
blocks_low_rank <- list(
  s1 = list(
    task = alignment_feature_block(
      matrix(rnorm(30), 3, 10), "task",
      feature_names = c("f1", "f2", "f3")
    )
  )
)

diag <- feature_block_diagnostics(
  blocks_low_rank$s1,
  convention = "right",
  include_singular_values = TRUE
)

diag$stacked
```

The `fraction_identified_rank` tells you what proportion of the transform
space is constrained. Values below 0.5 suggest significant rank deficiency.

### Computing the identified subspace

`identified_subspace_basis()` returns the orthonormal basis spanning the
identified dimensions:

```{r subspace-basis}
set.seed(4)
n <- 20  # ambient dimension
k <- 5   # correspondence dimension

# A rank-5 correspondence matrix in 20D
X <- matrix(rnorm(k * 10), k, 10)
embedding <- matrix(rnorm(n * k), n, k)
correspondence <- embedding %*% X  # 20 x 10, rank 5

sub <- identified_subspace_basis(correspondence, convention = "left")
sub$numeric_rank
sub$transform_dim
dim(sub$basis)  # 20 x 5 orthonormal basis
```

### Projecting to the subspace

Project data or transforms onto the identified subspace for stable inference:

```{r project-subspace}
# Full-space vector
v <- rnorm(n)

# Project to identified subspace
v_proj <- project_to_subspace(v, sub$basis, side = "left")
length(v_proj)  # still n-dimensional, but lies in the subspace
```

### Subspace coordinates

Convert to reduced-dimensional coordinates:

```{r subspace-coords}
z <- subspace_coordinates(v, sub$basis, side = "left")
length(z)  # k-dimensional
```

### Restricting operators to the subspace

If `Q` is a full-space operator, its restriction to the identified subspace
captures only the identifiable part:

```{r restrict-operator}
Q_full <- qr.Q(qr(matrix(rnorm(n^2), n, n)))  # 20x20 orthogonal
Q_restricted <- restrict_operator_to_subspace(Q_full, sub$basis)
dim(Q_restricted)  # 5x5
```

### Lifting back to full space

Convert a subspace operator back to the ambient space:

```{r lift-operator}
Q_lifted <- lift_operator_from_subspace(Q_restricted, sub$basis, fill = "identity")
dim(Q_lifted)  # 20x20

# Acts like Q_full in the subspace, like identity in the complement
z_test <- rnorm(sub$numeric_rank)
x_in_subspace <- sub$basis %*% z_test
max(abs(Q_lifted %*% x_in_subspace - Q_full %*% x_in_subspace))
```

## Part 3: Multi-stage pipelines

### Composing alignment models

Real-world pipelines often involve multiple alignment stages:

1. **Anatomical alignment** (e.g., spatial normalization to MNI space)
2. **Functional alignment** (e.g., Procrustes on task activations)
3. **Fine-tuning** (e.g., optimal transport refinement)

neuralign can compose these into a single model:

```{r compose-models}
set.seed(5)
p <- 10
n_obs <- 20

shared <- matrix(rnorm(p * n_obs), p, n_obs)

data_list <- lapply(1:3, function(i) {
  Q <- qr.Q(qr(matrix(rnorm(p^2), p)))
  Q %*% shared + 0.3 * matrix(rnorm(p * n_obs), p, n_obs)
})
names(data_list) <- paste0("s", 1:3)

# Stage 1: Procrustes with medoid reference
adat_stage1 <- AlignmentData(data_list, space = "demo-feature-space")
result1 <- fit_alignment(
  adat_stage1,
  method = "procrustes",
  reference = "medoid"
)

# Stage 2: Procrustes again with consensus (on already-aligned data)
aligned1 <- get_aligned(result1)
result2 <- fit_alignment(
  AlignmentData(
    aligned1,
    space = get_model(result1)@space_to
  ),
  method = "procrustes",
  reference = "consensus"
)

# Compose: apply stage 1 first, then stage 2
model1 <- get_model(result1)
model2 <- get_model(result2)
```

### Checking composition compatibility

Before composing, verify that the models are compatible:

```{r check-compose}
compat <- check_composition(model1, model2)
compat$compatible
compat$message
```

Here the input was assigned an explicit feature-space identity, and the second
stage inherited the first stage's output space. Composition fails closed when
that intermediate identity is absent or disagrees between stages.

```{r check-compose-contract, include=FALSE}
stopifnot(isTRUE(compat$compatible))
```

### Creating the composed model

```{r do-compose}
composed <- compose_alignment(model1, model2)
composed
```

The composed model contains a single transform per subject that is the
product of the two stages: `T_composed = T2 %*% T1`.

### Operator composition

You can also use the `%*%` operator:

```{r compose-operator, eval=FALSE}
composed <- model2 %*% model1  # model1 applied first, then model2
```

### Applying the composed model

```{r apply-composed}
# New observations for existing subjects
new_obs <- AlignmentData(lapply(data_list, function(m) {
  matrix(rnorm(p * 5), p, 5)  # 5 new timepoints per subject
}))

# Apply the full composed pipeline in one step
result_composed <- apply_alignment(composed, new_obs)
names(get_aligned(result_composed))
```

## Part 4: Block alignment report

The `block_alignment_report()` function ties everything together: harmonization,
identifiability diagnostics, and quality metrics in a single report.

```{r block-report}
set.seed(6)

blocks_by_subject <- list(
  s1 = list(
    task = alignment_feature_block(
      matrix(rnorm(40), 4, 10), "task",
      feature_names = paste0("f", 1:4), weight = 1
    ),
    anat = alignment_feature_block(
      matrix(rnorm(20), 2, 10), "anat",
      feature_names = c("thick", "curv"), weight = 0.5
    )
  ),
  s2 = list(
    task = alignment_feature_block(
      matrix(rnorm(40), 4, 10), "task",
      feature_names = paste0("f", 1:4), weight = 1
    ),
    anat = alignment_feature_block(
      matrix(rnorm(20), 2, 10), "anat",
      feature_names = c("thick", "curv"), weight = 0.5
    )
  )
)

report <- block_alignment_report(blocks_by_subject, convention = "right")
report
```

When you also have aligned data, include it for quality metrics:

```{r block-report-quality}
# Build features and align
features <- build_alignment_features(blocks_by_subject)
result <- fit_alignment(
  AlignmentData(features$matrices),
  method = "procrustes"
)

report_with_quality <- block_alignment_report(
  blocks_by_subject,
  aligned = result,
  quality_metrics = c("correlation")
)
report_with_quality
```

## Summary

| Component | Key Functions |
|-----------|---------------|
| **Feature blocks** | `alignment_feature_block()`, `harmonize_feature_blocks()`, `stack_feature_blocks()`, `build_alignment_features()` |
| **Identifiability** | `feature_block_diagnostics()`, `identified_subspace_basis()` |
| **Subspace operations** | `project_to_subspace()`, `subspace_coordinates()`, `restrict_operator_to_subspace()`, `lift_operator_from_subspace()` |
| **Model composition** | `compose_alignment()`, `check_composition()` |
| **Diagnostics** | `block_alignment_report()` |

### When to use what

- **Few, reliable correspondence features**: Use feature blocks with
  intersection harmonization. Check identifiability; if low, consider
  restricting to the identified subspace.
- **Sparse, heterogeneous features**: Use union-fill harmonization. Track
  coverage per subject.
- **Multiple alignment stages**: Compose models rather than running them
  independently. This enables single-step application to new data and correct
  provenance tracking.
- **Rank-deficient correspondence**: Use subspace operations to restrict
  inference to identified dimensions. This prevents overfitting to
  unconstrained transform directions.
