---
title: "Choosing Anchors and Guidance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Anchors and Guidance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Core idea: you need a shared axis

All alignment methods require **some notion of correspondence across subjects**. In practice,
neuralign supports three common regimes:

1. **Shared observation labels** (classic task alignment): subjects share a condition/trial/time axis.
2. **Sparse label overlap**: subjects have partially overlapping label sets, but the overlap graph is connected.
3. **No shared labels** (e.g., resting-state embeddings): you manufacture a comparable anchor space via anatomy/geometry guidance.

This vignette shows the primitives neuralign provides for each regime.

```{r setup}
library(neuralign)
```

## 1) Shared labels: Procrustes

If all subjects share the same observation axis (e.g., timepoints or trial labels), use the built-in
Procrustes aligner:

```{r shared-labels}
set.seed(1)
X <- matrix(rnorm(50), 5, 10)
Q <- qr.Q(qr(matrix(rnorm(25), 5, 5)))
if (det(Q) < 0) Q[, 1] <- -Q[, 1]

adat <- AlignmentData(list(
  s1 = X,
  s2 = Q %*% X
))

res <- fit_alignment(adat, method = "procrustes", reference = "s1")
get_quality(res)
```

## 2) Sparse overlap: graph-synchronized Procrustes

When subjects have **different subsets of labels** (possibly empty global intersection), but the
overlap graph is connected, use `"procrustes_graph"`. Supply per-subject `obs_labels`:

```{r sparse-overlap}
set.seed(2)
d <- 4
labels <- paste0("stim", 1:8)
Z <- matrix(rnorm(d * length(labels)), d, length(labels))
rand_rot <- function(d) {
  q <- qr.Q(qr(matrix(rnorm(d * d), d, d)))
  if (det(q) < 0) q[, 1] <- -q[, 1]
  q
}
A1 <- rand_rot(d)
A2 <- rand_rot(d)
A3 <- rand_rot(d)

adat <- AlignmentData(
  data = list(
    s1 = A1 %*% Z[, 1:4, drop = FALSE],
    s2 = A2 %*% Z[, 1:8, drop = FALSE],
    s3 = A3 %*% Z[, 5:8, drop = FALSE]
  ),
  obs_labels = list(
    s1 = labels[1:4],
    s2 = labels[1:8],
    s3 = labels[5:8]
  )
)

res <- fit_alignment(
  adat,
  method = "procrustes_graph",
  reference = "s1",
  min_overlap = d
)

get_quality(res)
```

The result aligns each subject into a **globally consistent** reference space, without requiring a
full label intersection.

## 3) Shared feature bases ("betas-as-features"): feature blocks

If subjects share a **named feature basis** (e.g., encoding-model betas), you can treat the feature
axis as the shared coordinate system. neuralign provides block helpers for harmonizing and stacking
named feature matrices:

```{r feature-blocks}
set.seed(3)

b1 <- alignment_feature_block(
  x = matrix(rnorm(6), 3, 2),
  name = "betas",
  weight = 1,
  feature_names = c("f1", "f2", "f3")
)

b2 <- alignment_feature_block(
  x = matrix(rnorm(6), 3, 2),
  name = "betas",
  weight = 1,
  feature_names = c("f3", "f2", "f1") # different order
)

blocks <- harmonize_feature_blocks(list(
  s1 = list(betas = b1),
  s2 = list(betas = b2)
))

X1 <- stack_feature_blocks(blocks$s1)
X2 <- stack_feature_blocks(blocks$s2)
dim(X1)
```

Downstream packages can use these matrices as standardized inputs to alignment.

### Feature blocks in downstream workflows

Many downstream workflows build a per-subject correspondence matrix `F_s` by stacking multiple
signals (task-derived features, anatomical priors, learned embeddings, etc.). neuralign’s
feature blocks standardize the *plumbing* across packages:

- represent each signal as a named block with a weight and optional feature names,
- harmonize blocks across subjects by intersecting/reordering feature names,
- stack blocks with a single, explicit weight convention.

Blocks may also carry optional `meta` fields (opaque to neuralign) for provenance and tags.
See the fMRI profile vignette for an example naming scheme (optional):

```r
vignette("fmri_profile", package = "neuralign")
```

#### Orientation: left vs right Procrustes conventions

neuralign’s alignment API uses a **left-multiply** convention (operators act on rows of
`features × observations` matrices). Some workflows prefer a **right-multiply** basis-rotation
view on a correspondence matrix:

- right convention: `F_s %*% O_s ≈ F_ref` where `F_s` is
  `(correspondence_rows × K)` and `O_s` is `(K × K)`.
- left convention (neuralign style): `Q_s %*% X_s ≈ X_ref` where `X_s = t(F_s)` is
  `(K × correspondence_rows)` and `Q_s = t(O_s)`.

neuralign exposes both conventions via `procrustes_rotation()`:

```{r procrustes-conventions}
set.seed(10)
P <- 6
K <- 3
O_true <- qr.Q(qr(matrix(rnorm(K * K), K, K)))
if (det(O_true) < 0) O_true[, 1] <- -O_true[, 1]

F_ref <- matrix(rnorm(P * K), P, K)
F_s <- F_ref %*% t(O_true) # so F_s %*% O_true = F_ref

rot_right <- procrustes_rotation(F_s, F_ref, convention = "right")
rot_left <- procrustes_rotation(t(F_s), t(F_ref), convention = "left")

max(abs(rot_right$Q - O_true))
max(abs(t(rot_left$Q) - O_true))
```

#### Weight semantics (important!)

`stack_feature_blocks()` scales each block by `sqrt(weight)` so weights behave naturally in a
sum-of-squares objective. If a legacy pipeline multiplied blocks by a linear weight `w`
directly, an equivalent neuralign weight is `weight = w^2`.

If a legacy pipeline multiplied blocks by a linear weight `w` directly, an equivalent
neuralign weight is `weight = w^2`.

To quantify how much of the transform space is identified by your correspondence signal(s), use
`feature_block_diagnostics()`:

```{r feature-block-diagnostics}
feature_block_diagnostics(blocks$s1, convention = "right")$stacked
```

#### Minimal recipe

1. Build per-subject block lists with `alignment_feature_block()` (set `feature_names` when you
   want harmonization by feature identity).
2. Call `harmonize_feature_blocks()` to intersect and reorder features within each block.
3. Call `stack_feature_blocks()` to obtain `F_s` (a single stacked matrix).
4. Use `procrustes_rotation()` with the convention appropriate to your representation.

If your correspondence matrix is naturally stored as `(K × D)` (basis/components × correspondence),
use `t()` when needed so that rows correspond to named correspondence features.

If a block has **no shared features** across subjects (e.g., empty intersection), it will be
dropped by harmonization; if *all* correspondence signals vanish, you must switch regimes (see
Section 2 on sparse overlap and Section 4–5 on anatomical/geometry guidance).

#### Sparse overlap within a correspondence signal

If a correspondence signal has **no global intersection** (e.g., each subject saw a different
subset of stimuli), do not force an intersection via `harmonize_feature_blocks()`. If the overlap
graph is connected, use `"procrustes_graph"` on the raw per-subject matrices with per-subject
`obs_labels`. You can still recover a right-multiply rotation by transposing the left-multiply
transform:

```{r procrustes-graph-right-rotation}
set.seed(11)
K <- 3
stim <- paste0("stim", 1:8)
Z <- matrix(rnorm(K * length(stim)), K, length(stim))
A <- qr.Q(qr(matrix(rnorm(K * K), K, K)))
if (det(A) < 0) A[, 1] <- -A[, 1]

B1 <- A %*% Z[, 1:4, drop = FALSE]
B2 <- A %*% Z[, 1:8, drop = FALSE]
B3 <- A %*% Z[, 5:8, drop = FALSE]

adat <- AlignmentData(
  data = list(s1 = B1, s2 = B2, s3 = B3),
  obs_labels = list(s1 = stim[1:4], s2 = stim[1:8], s3 = stim[5:8])
)

res <- fit_alignment(adat, method = "procrustes_graph", reference = "s1", min_overlap = K)
Q2 <- get_transform(get_model(res), "s2") # left-multiply transform
O2 <- t(Q2)                               # right-multiply rotation
dim(O2)
```

## 4) ROI/parcel anchors: projector guidance

If you have subject-specific ROI labels (even outside template space), you can build a projector
`P_s` of shape `(anchors x features)` and store it as guidance:

```{r roi-guidance}
set.seed(4)
data <- AlignmentData(list(
  s1 = matrix(rnorm(12), 3, 4),
  s2 = matrix(rnorm(15), 3, 5)
))

roi <- list(
  s1 = c("V1", "V1", "MT"),
  s2 = c("V1", "V1", "V1")
)

data2 <- set_roi_guidance(data, roi_by_subject = roi, anchor_policy = "union")
get_guidance(data2, type = "projector")
```

Projectors can be used to create shared anchor summaries (e.g., `P_s %*% X_s`) or as anatomical costs
in methods that support guided alignment.

## 5) Intrinsic geometry: coordinate guidance

When ROI labels are unavailable or unreliable, you can compute smooth intrinsic coordinates from a
subject-specific adjacency graph (e.g., mesh or voxel neighborhood graph):

```{r intrinsic-geometry}
path_adj <- function(n) {
  A <- matrix(0, n, n)
  for (i in seq_len(n - 1L)) {
    A[i, i + 1L] <- 1
    A[i + 1L, i] <- 1
  }
  A
}

data <- AlignmentData(list(
  s1 = matrix(rnorm(20), 5, 4),
  s2 = matrix(rnorm(24), 6, 4)
))

adj <- list(
  s1 = path_adj(5),
  s2 = path_adj(6)
)

data2 <- set_intrinsic_geometry_guidance(
  data,
  adjacency_by_subject = adj,
  k = 2,
  method = "eigen"
)

get_guidance(data2, type = "coords")
```

These coordinates are comparable only up to sign/rotation ambiguities and are intended as priors for
guided alignment objectives (e.g., fused costs for OT/GW-style methods).
