---
title: "Cross-Validation and Leakage Prevention"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-Validation and Leakage Prevention}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why cross-validation matters for alignment

Neural alignment learns a transformation for each subject from training data.
If you evaluate the alignment on the same data used to fit it, you will
overestimate quality: the transforms have been optimized to make *those
particular observations* look similar. Proper cross-validation holds out data
that the fitting procedure never sees, giving an honest estimate of
generalization.

neuralign provides two orthogonal CV axes, built-in leakage detection, and
careful anchor-space tracking---all integrated into the standard
`fit_alignment()` interface.

```{r setup}
library(neuralign)
```

## Subject-axis cross-validation

Subject-axis CV answers the question: *"How well does this alignment
generalize to a new subject?"*

### Leave-one-subject-out (LOSO)

Each subject is held out once. The model is trained on the remaining subjects,
and the held-out subject's transform is estimated from the training model.

```{r loso}
set.seed(1)
n <- 30; p <- 5
shared <- matrix(rnorm(p * n), p, n)

data_list <- lapply(1:4, function(i) {
  Q <- qr.Q(qr(matrix(rnorm(p^2), p)))
  Q %*% shared + 0.3 * matrix(rnorm(p * n), p, n)
})
names(data_list) <- paste0("s", 1:4)
adat <- AlignmentData(data_list)

result_loso <- fit_alignment(adat, method = "procrustes", cv = "loso")
get_cv_info(result_loso)
```

With 4 subjects, LOSO produces 4 folds. Each subject's aligned output was
produced by a model that never saw that subject's data during training.

### K-fold cross-validation

When you have many subjects, LOSO becomes expensive. K-fold partitions
subjects into *k* groups:

```{r kfold}
set.seed(2)
data_list_large <- lapply(1:12, function(i) {
  Q <- qr.Q(qr(matrix(rnorm(p^2), p)))
  Q %*% shared + 0.3 * matrix(rnorm(p * n), p, n)
})
names(data_list_large) <- paste0("s", sprintf("%02d", 1:12))
adat_large <- AlignmentData(data_list_large)

result_kf <- fit_alignment(
  adat_large,
  method = "procrustes",
  cv = "kfold",
  cv_folds = 4
)
get_cv_info(result_kf)
```

### Creating fold specifications manually

For full control, use `create_cv_folds()` and pass the result to
`fit_alignment()`:

```{r custom-folds}
folds <- create_cv_folds(adat, method = "kfold", k = 2, seed = 99)
folds$assignments

result_custom <- fit_alignment(adat, method = "procrustes", cv_folds = folds)
get_cv_info(result_custom)$n_folds
```

## Observation-axis cross-validation

Subject-axis CV evaluates generalization to *new subjects*. But in many
workflows, you need to evaluate generalization across *observations*
(timepoints, runs, trials) while keeping subjects fixed. This is especially
important when the correspondence signal (e.g., task betas) is estimated from
the same observations you later evaluate---using the same data for both
creates circular analysis.

### Run-wise folds

If your experiment has natural blocks (runs, sessions, conditions), each block
can serve as one fold:

```{r obs-run-folds}
set.seed(3)
n_obs <- 12
runs <- rep(paste0("run", 1:3), each = 4)
obs_labels <- paste0("t", seq_len(n_obs))

shared2 <- matrix(rnorm(p * n_obs), p, n_obs)
data_list2 <- lapply(1:3, function(i) {
  Q <- qr.Q(qr(matrix(rnorm(p^2), p)))
  Q %*% shared2 + 0.3 * matrix(rnorm(p * n_obs), p, n_obs)
})
names(data_list2) <- paste0("s", 1:3)
adat2 <- AlignmentData(data_list2, obs_labels = obs_labels)

obs_folds <- create_obs_folds(runs, method = "run")
obs_folds$n_folds
obs_folds$fold_ids
```

### Blocked-time folds

When there are no natural run boundaries, create contiguous temporal blocks:

```{r obs-blocked}
obs_folds_time <- create_obs_folds(n_obs, method = "blocked_time", k = 3)
obs_folds_time$n_folds
```

### Guard TRs: preventing temporal leakage

In fMRI data, adjacent timepoints are autocorrelated. Holding out timepoint
*t* but training on *t-1* and *t+1* leaks information. Guard TRs exclude a
buffer of timepoints around each held-out block:

```{r guard-tr}
obs_folds_guarded <- create_obs_folds(
  20,
  method = "blocked_time",
  k = 4,
  guard_tr = 2  # exclude 2 TRs before/after each test block
)
# Training set for fold 1 excludes the test block PLUS 2 adjacent TRs on each side
```

### Fitting with observation-axis folds

Pass observation-axis fold specs directly to `fit_alignment()`:

```{r obs-fit}
cf <- run_obs_crossfit_from_data(
  adat2,
  obs_folds,
  method = "procrustes",
  reference = "s1",
  anchor_policy = "common_or_error"
)

cf$anchor_common
names(cf$aligned_test_by_fold)
```

## Anchor semantics: common vs. fold-specific references

When the reference is **data-driven** (e.g., `"medoid"` or `"consensus"`),
each CV fold selects its own reference from the training data. This means
fold outputs may live in *different coordinate systems*---they are
**fold-specific anchors**.

### Detecting anchor type

```{r anchor-detect}
# Fixed reference -> always common anchor
cf_fixed <- run_obs_crossfit_from_data(
  adat2,
  obs_folds,
  method = "procrustes",
  reference = "s1",
  anchor_policy = "common_or_error"
)
cf_fixed$anchor_common
```

```{r anchor-medoid}
# Data-driven reference -> may produce fold-specific anchors
cf_med <- suppressWarnings(run_obs_crossfit_from_data(
  adat2,
  obs_folds,
  method = "procrustes",
  reference = "medoid",
  anchor_policy = "fold_specific_ok"
))
cf_med$anchor_common
```

### Anchor policies

| Policy | Behavior |
|--------|----------|
| `"common_or_error"` | Error if reference is data-driven (enforces a single anchor space) |
| `"fold_specific_ok"` | Allow fold-specific anchors (mark `anchor_common = FALSE`) |
| `"map_to_template"` | Map fold-specific anchors to an external template via Procrustes |

### When anchor matters

- **`anchor_common = TRUE`**: safe to concatenate or average across folds.
- **`anchor_common = FALSE`**: each fold's output lives in a different space. Do
  not aggregate directly. Either use a fixed reference or map to a template.

## Leakage detection

neuralign tracks which subjects were used to train a model and warns when you
accidentally evaluate on training data.

### Automatic warning in apply_alignment()

```{r leakage-warning}
# Fit on all subjects
model <- get_model(fit_alignment(adat, method = "procrustes"))

# Apply to training subjects -> triggers leakage warning
new_result <- apply_alignment(model, adat, warn_leakage = TRUE)
```

### Risk assessment

For a comprehensive leakage audit, use `assess_leakage_risk()`:

```{r leakage-assess}
risk <- assess_leakage_risk(
  result_loso,
  apply_subjects = adat@subjects
)
risk$overall_risk
risk$issues
risk$recommendations
```

### Observation-level leakage

`check_obs_leakage()` detects overlapping observation indices in observation-axis
CV setups:

```{r obs-leakage, eval=FALSE}
check_obs_leakage(
  train_obs = list(s1 = 1:10, s2 = 1:10),
  test_obs = list(s1 = 8:15, s2 = 8:15),  # overlaps with train!
  action = "error"
)
```

## Putting it together: a safe evaluation workflow

```{r safe-workflow}
set.seed(42)

# 1. Create data
adat_eval <- AlignmentData(data_list)

# 2. Fit with LOSO cross-validation
result <- fit_alignment(adat_eval, method = "procrustes", cv = "loso")

# 3. Check CV metadata
cv_info <- get_cv_info(result)
cat("CV method:", cv_info$method, "\n")
cat("Folds:", cv_info$n_folds, "\n")

# 4. Assess quality (aligned data is already cross-validated)
quality <- alignment_quality(result, original = adat_eval, metrics = "correlation")
cat("Aligned correlation:", quality$mean_pairwise_correlation, "\n")
cat("Original correlation:", quality$original_mean_correlation, "\n")
cat("Improvement:", quality$correlation_improvement, "\n")
```

## Summary of CV functions

| Function | Purpose |
|----------|---------|
| `create_cv_folds()` | Create subject-axis fold specifications (LOSO, k-fold, stratified) |
| `create_obs_folds()` | Create observation-axis fold specifications (run-wise, blocked-time) |
| `fit_alignment(..., cv = "loso")` | Fit with built-in subject-axis CV |
| `run_obs_crossfit_from_data()` | Observation-axis crossfit from AlignmentData |
| `run_obs_crossfit_alignment()` | Observation-axis crossfit from pre-split fold data |
| `validate_cv_setup()` | Validate fold structure (disjointness, reference safety) |
| `check_obs_leakage()` | Detect overlapping train/test observations |
| `assess_leakage_risk()` | Holistic leakage risk assessment |
| `has_common_anchor()` | Check whether CV result has a single anchor space |
| `validate_common_anchor()` | Error/warn if anchor is fold-specific |
