---
title: "Extending neuralign: Custom Alignment Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending neuralign: Custom Alignment Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

neuralign is designed to be extended. Any R package (or even a script) can
register a new alignment method that then works seamlessly with
`fit_alignment()`, `apply_alignment()`, cross-validation, composition, quality
assessment, and serialization. This vignette walks through the process.

```{r setup}
library(neuralign)
```

## The aligner contract

Every alignment method must provide a **fit function** and optionally an
**apply function**. neuralign calls these at the right time and feeds them
standardized inputs.

### Fit function signature

```r
fit_fn <- function(data, reference, train_idx = NULL, ...) {
  # data: AlignmentData object
  # reference: subject ID (character), "consensus", or a matrix
  # train_idx: integer indices of subjects to use (NULL = all)
  # ...: method-specific parameters from the user

  list(
    transforms = list(...),       # Named list: subject_id -> operator matrix
    reference_data = matrix(...), # The actual reference data used
    space_from = data@space,      # Source space (can be NULL)
    space_to = data@space,        # Target space (can be NULL)
    method_state = list(...)      # Anything needed by apply_fn later
  )
}
```

The fit function receives an `AlignmentData` object containing multi-subject
data, a reference specification, and optional training indices (for
cross-validation). It returns per-subject operators that transform each
subject's data into the reference space.

**Key requirements:**

- `transforms` must be a named list where each value is a matrix (or
  Matrix object) that left-multiplies the subject's data:
  `aligned = transform %*% subject_data`.
- Transform dimensions: `(target_features × source_features)`.
- `reference_data` is the actual matrix/template used as the alignment target.

### Apply function signature (optional)

```r
apply_fn <- function(fit_result, new_data, ...) {
  # fit_result: the list returned by fit_fn (transforms, reference_data,
  #             method_state, etc.)
  # new_data: AlignmentData with a SINGLE new subject
  # ...: method-specific parameters

  list(
    transforms = list(subject_id = operator_matrix),
    reference_data = fit_result$reference_data,
    space_from = fit_result$space_from,
    space_to = fit_result$space_to,
    method_state = fit_result$method_state
  )
}
```

If you do not provide an `apply_fn`, neuralign will re-call `fit_fn` with the
new subject appended to the data. For methods where this is efficient (or the
only option), you can omit `apply_fn`.

## Declaring capabilities

Each method declares what it can do and what it requires via a **capabilities
list**. This enables neuralign to validate data before fitting, route
cross-validation correctly, and provide accurate diagnostics.

```{r capabilities-example}
my_capabilities <- list(
  # Cross-validation support
  supports_cv = TRUE,
  cv_axes = c("subject"),         # Which axes: "subject", "observation", or both

  # Data requirements
  needs_geometry = FALSE,          # Requires adjacency matrix?
  needs_design = FALSE,            # Requires design/task structure?
  needs_guidance = FALSE,          # Requires guidance channels?
  guidance_types = character(0),   # Required guidance types if needs_guidance

  # Data shape requirements
  requires_shared_features = TRUE, # All subjects must have same feature count?
  requires_shared_observations = TRUE, # All subjects must have same obs count?

  # Transform properties
  returns = "operator",            # "operator" (matrix) or "embedding"
  transform_type = "linear",       # "orthogonal", "linear", "ot", "permutation"
  returns_invertible = FALSE,      # Is the transform exactly invertible?
  mass_preserving = FALSE,         # For OT: does transport preserve mass?

  # Application capabilities
  supports_new_subject = TRUE,     # Can compute transform for unseen subject?
  supports_new_data = TRUE,        # Can apply existing transform to new data?

  # Reference options
  reference_types = c("subject", "consensus")
)
```

**Important capability flags:**

- `returns = "operator"` means transforms are matrices that left-multiply data.
  This is currently the only supported mode.
- `transform_type` informs inversion strategy: `"orthogonal"` transforms use
  transpose for inversion, others may need pseudoinverse.
- `reference_types` lists which reference specifications the method supports.
  neuralign validates the user's choice against this list.

## Example: a scaling aligner

Let's implement a simple alignment method that scales each subject's data to
match the reference subject's variance. This is not a realistic alignment
method, but it illustrates the full registration flow.

### Step 1: Write the fit function

```{r scaling-fit}
.scaling_fit <- function(data, reference, train_idx = NULL, ...) {
  # Determine which subjects to train on
  if (!is.null(train_idx)) {
    train_data <- data[train_idx]
  } else {
    train_data <- data
  }

  # Resolve reference
  if (is.character(reference) && reference %in% data@subjects) {
    ref_data <- get_subject_data(data, reference)
  } else if (is.character(reference) && reference == "consensus") {
    ref_data <- Reduce("+", get_data_list(train_data)) / length(train_data)
  } else if (is.matrix(reference)) {
    ref_data <- reference
  } else {
    stop("Unsupported reference type")
  }

  # Compute per-feature variance of the reference
  ref_var <- apply(ref_data, 1, var)
  ref_var[ref_var < 1e-12] <- 1  # avoid division by zero

  # For each subject, compute a diagonal scaling transform
  transforms <- list()
  for (subj in data@subjects) {
    subj_data <- get_subject_data(data, subj)
    subj_var <- apply(subj_data, 1, var)
    subj_var[subj_var < 1e-12] <- 1

    scale_factors <- sqrt(ref_var / subj_var)
    transforms[[subj]] <- diag(scale_factors)
  }

  list(
    transforms = transforms,
    reference_data = ref_data,
    space_from = data@space,
    space_to = data@space,
    method_state = list(ref_var = ref_var)
  )
}
```

### Step 2: Define capabilities

```{r scaling-caps}
.scaling_capabilities <- list(
  supports_cv = TRUE,
  cv_axes = c("subject"),
  needs_geometry = FALSE,
  needs_design = FALSE,
  needs_guidance = FALSE,
  requires_shared_features = TRUE,
  requires_shared_observations = FALSE,
  returns = "operator",
  transform_type = "linear",
  returns_invertible = TRUE,
  mass_preserving = FALSE,
  supports_new_subject = TRUE,
  supports_new_data = TRUE,
  reference_types = c("subject", "consensus")
)
```

### Step 3: Register

```{r scaling-register}
register_aligner(
  name = "variance_scaling",
  fit_fn = .scaling_fit,
  apply_fn = NULL,  # will re-call fit_fn for new subjects
  capabilities = .scaling_capabilities,
  package = "neuralign",
  description = "Diagonal variance scaling",
  version = "0.1.0"
)

# Verify registration
"variance_scaling" %in% available_aligners()
```

### Step 4: Use it

```{r scaling-use}
set.seed(10)
data_list <- list(
  s1 = matrix(rnorm(20), 5, 4),
  s2 = matrix(rnorm(20) * 3, 5, 4),  # 3x larger variance
  s3 = matrix(rnorm(20) * 0.5, 5, 4) # 0.5x smaller variance
)
adat <- AlignmentData(data_list)

result <- fit_alignment(adat, method = "variance_scaling", reference = "s1")
result
get_quality(result)
```

### Step 5: Cross-validation works automatically

```{r scaling-cv}
result_cv <- fit_alignment(
  adat,
  method = "variance_scaling",
  reference = "consensus",
  cv = "loso"
)
get_cv_info(result_cv)$n_folds
```

### Step 6: Applying to new data works automatically

```{r scaling-apply}
new_data <- AlignmentData(list(s4 = matrix(rnorm(20) * 2, 5, 4)))
applied <- apply_alignment(result, new_data)
names(get_aligned(applied))
```

### Cleanup

```{r scaling-cleanup}
unregister_aligner("variance_scaling")
```

## Contract validation

neuralign validates your method at registration time. If your `fit_fn` or
`apply_fn` signatures don't match the contract, registration will fail:

```{r contract-validation, error=TRUE}
# This will fail: fit_fn missing required 'data' parameter
tryCatch(
  register_aligner(
    name = "bad_method",
    fit_fn = function(x, y) list(),  # wrong signature!
    capabilities = list()
  ),
  error = function(e) message("Registration failed: ", e$message)
)
```

Use `validate_aligner_contract()` to check your functions before registering:

```{r validate-contract, error=TRUE}
tryCatch(
  validate_aligner_contract(
    name = "test",
    fit_fn = function(data, reference, ...) list(),
    capabilities = list(returns = "invalid_value")
  ),
  error = function(e) message("Validation failed: ", e$message)
)
```

## Registering from an external package

In practice, you register your method in your package's `.onLoad()` hook so it
becomes available as soon as the package is loaded:

```r
# In mypackage/R/zzz.R:
.onLoad <- function(libname, pkgname) {
  if (requireNamespace("neuralign", quietly = TRUE)) {
    neuralign::register_aligner(
      name = "my_method",
      fit_fn = .my_method_fit,
      apply_fn = .my_method_apply,
      capabilities = .my_method_caps,
      package = pkgname,
      description = "My alignment method",
      version = as.character(utils::packageVersion(pkgname))
    )
  }
}
```

Users then do:

```r
library(neuralign)
library(mypackage)  # triggers registration

fit_alignment(data, method = "my_method")
```

### Auto-loading

neuralign also supports **auto-loading**: if a user requests a method that
isn't registered but neuralign knows which package provides it, it will
attempt to load that package. To add your package to the known-method
mapping, contact the neuralign maintainers or submit a PR.

## API versioning

neuralign declares `NEURALIGN_ALIGNER_API_VERSION` (currently `r NEURALIGN_ALIGNER_API_VERSION`).
Your registration can declare which API version it implements. If a future
neuralign release bumps the API version, older registrations will still work
but may trigger compatibility warnings.

```{r api-version}
NEURALIGN_ALIGNER_API_VERSION
```

## Querying the registry

```{r registry-queries}
# List all registered methods
available_aligners()

# With details
available_aligners(details = TRUE)

# Get full entry for a method
entry <- get_aligner("procrustes")
names(entry)

# Get just capabilities
caps <- aligner_capabilities("procrustes")
caps$transform_type
caps$returns_invertible
```

## Summary

To register a custom alignment method:

1. Write a `fit_fn(data, reference, train_idx, ...)` that returns a list with
   `transforms`, `reference_data`, and optionally `space_from`, `space_to`,
   `method_state`.
2. Optionally write an `apply_fn(fit_result, new_data, ...)` for efficient
   application to new subjects.
3. Declare capabilities (CV support, data requirements, transform properties).
4. Call `register_aligner()` with your functions and capabilities.
5. Your method now works with `fit_alignment()`, `apply_alignment()`,
   cross-validation, composition, quality assessment, and serialization.
