---
title: "fMRI Alignment Profile (Recipes)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fMRI Alignment Profile (Recipes)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is an **optional recipe layer** for fMRI packages building on top of
neuralign. It provides naming conventions and workflow guidance, but the neuralign
core remains **domain-agnostic**: it operates on matrices + metadata and does not
encode fMRI-specific feature extraction rules.

```{r setup}
library(neuralign)
```

## 1) fMRI profile: what it standardizes (and what it does not)

**neuralign core owns:**

- block representation (`alignment_feature_block()`),
- harmonization/stacking (`harmonize_feature_blocks()`, `stack_feature_blocks()`),
- alignment backends (`fit_alignment()`, `procrustes_rotation()`, `"procrustes_graph"`),
- crossfit orchestration helpers (`run_obs_crossfit_alignment()`).

**Client packages (fMRI-specific) own:**

- how you compute task betas/contrasts/priors from native objects,
- what to exclude (nuisance regressors, degenerate columns, etc.),
- how you apply transforms to package-specific model state (e.g., components, maps),
- how you split folds to avoid leakage for supervised signals.

## 2) Block vocabulary (optional)

These are *just strings*; neuralign does not interpret them. A common vocabulary helps
interoperability and diagnostics:

- `task_beta`: GLM betas for selected design columns.
- `contrast_beta`: contrast betas (linear combinations of betas).
- `roi_fingerprint`: ROI-level summaries of functional patterns used as correspondence.
- `anat_feature`: coarse anatomy/geometry-derived correspondence features (guidance, not the target).
- `connectivity_fingerprint`: optional connectivity summaries (if used as correspondence).

## 3) Recommended `meta` tags (optional)

Blocks can carry a `meta` list for provenance and safety annotations. Suggested keys:

- `meta$source_type`: `"supervised"`, `"unsupervised"`, or `"prior"`
- `meta$requires_independence`: `TRUE/FALSE` (i.e., must be cross-fit for unbiased evaluation)

```{r meta-tags}
b <- alignment_feature_block(
  x = matrix(rnorm(12), 3, 4),
  name = "task_beta",
  weight = 1,
  feature_names = c("reg1", "reg2", "reg3"),
  meta = list(source_type = "supervised", requires_independence = TRUE)
)
b$meta
```

## 4) Basis alignment pattern (right-multiply view)

Many fMRI workflows align an ambiguous basis (e.g., latent components). A convenient
representation is a correspondence matrix `F_s` of shape:

- rows: correspondence features (regressors, contrasts, ROI features, etc.)
- columns: basis/components `K`

Then estimate `O_s` such that `F_s %*% O_s ≈ F_ref`. neuralign supports this via
`procrustes_rotation(convention="right")`.

```{r basis-alignment}
set.seed(1)
P <- 6
K <- 3

O_true <- qr.Q(qr(matrix(rnorm(K * K), K, K)))
if (det(O_true) < 0) O_true[, 1] <- -O_true[, 1]

F_ref <- matrix(rnorm(P * K), P, K)
F_s <- F_ref %*% t(O_true)

rot <- procrustes_rotation(F_s, F_ref, convention = "right")
max(abs(rot$Q - O_true))
```

## 5) Hybrid correspondence via feature blocks

Blocks let you combine signals with explicit weights:

```{r blocks-hybrid}
set.seed(2)

task <- alignment_feature_block(
  x = matrix(rnorm(12), 3, 4),
  name = "task_beta",
  weight = 1,
  feature_names = c("reg1", "reg2", "reg3"),
  meta = list(source_type = "supervised", requires_independence = TRUE)
)

anat <- alignment_feature_block(
  x = matrix(rnorm(8), 2, 4),
  name = "anat_feature",
  weight = 0.25,
  feature_names = c("roi:V1", "roi:MT"),
  meta = list(source_type = "prior", requires_independence = FALSE)
)

blocks <- harmonize_feature_blocks(list(
  s1 = list(task_beta = task, anat_feature = anat),
  s2 = list(task_beta = task, anat_feature = anat)
))

F1 <- stack_feature_blocks(blocks$s1)
feature_block_diagnostics(blocks$s1, convention = "right")$stacked
dim(F1)
```

## 6) Sparse overlap (subjects see different subsets)

When there is no global intersection of correspondence features (e.g., different subsets
of stimuli/conditions), but the overlap graph is connected, use `"procrustes_graph"` with
per-subject `obs_labels` rather than forcing a global intersection.

## 7) Observation-axis crossfit orchestration

When supervised blocks require independence (e.g., task-derived betas), compute them per fold
in the client package, then hand fold-wise matrices to neuralign:

```{r obs-crossfit}
set.seed(3)
d <- 3
labels <- paste0("run", 1:6)
Z <- matrix(rnorm(d * length(labels)), d, length(labels))

train <- list(
  f1 = list(s1 = Z[, 1:3, drop = FALSE], s2 = Z[, 1:3, drop = FALSE]),
  f2 = list(s1 = Z[, 4:6, drop = FALSE], s2 = Z[, 4:6, drop = FALSE])
)
test <- list(
  f1 = list(s1 = Z[, 4:6, drop = FALSE], s2 = Z[, 4:6, drop = FALSE]),
  f2 = list(s1 = Z[, 1:3, drop = FALSE], s2 = Z[, 1:3, drop = FALSE])
)

cf <- run_obs_crossfit_alignment(
  train_data_by_fold = train,
  test_data_by_fold = test,
  method = "procrustes",
  reference = "s1"
)
cf$anchor_common
names(cf$aligned_test_by_fold)
```

