---
title: "Introduction to neuralign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to neuralign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is neuralign?

neuralign provides a **unified interface** for neural alignment methods. It
standardizes how alignment models are fit, applied, composed, cross-validated,
and persisted---regardless of which algorithm or package implements the actual
computation.

### Architecture at a glance

```
  Raw data (per subject)
       |
       v
  AlignmentData           <-- S4 container: features x observations matrices
       |
       v
  fit_alignment()         <-- dispatches to registered aligner
       |
       v
  AlignmentResult         <-- holds model + aligned data + quality + CV info
       |
       +---> AlignmentModel   <-- per-subject transforms + reference + provenance
       +---> aligned data     <-- transformed matrices
       +---> quality metrics  <-- correlation, ISC, etc.
```

### Three core objects

| Object | Role | Key accessors |
|--------|------|---------------|
| `AlignmentData` | Input container: named list of `features x observations` matrices, plus optional metadata (design, geometry, space) | `get_data_list()`, `get_subject_data()` |
| `AlignmentModel` | Fitted model: per-subject transforms, reference data, method, provenance | `get_transform()`, `get_reference()` |
| `AlignmentResult` | Full result: wraps model + aligned data + quality + CV info | `get_model()`, `get_aligned()`, `get_quality()`, `get_cv_info()` |

```{r setup}
library(neuralign)
```

## Quick example

```{r quick-example}
set.seed(42)
n_features <- 20
n_obs <- 30

# Shared signal, rotated differently for each subject
shared <- matrix(rnorm(n_features * n_obs), n_features, n_obs)

data_list <- lapply(1:4, function(i) {
  Q <- qr.Q(qr(matrix(rnorm(n_features^2), n_features)))
  Q %*% shared + 0.3 * matrix(rnorm(n_features * n_obs), n_features, n_obs)
})
names(data_list) <- paste0("sub-", sprintf("%02d", 1:4))

# Wrap, fit, inspect
adat <- AlignmentData(data_list)
result <- fit_alignment(adat, method = "procrustes", reference = "consensus")
result
```

```{r quality}
get_quality(result)
```

```{r apply-new}
# Apply to a new (fifth) subject
new_sub <- list("sub-05" = {
  Q <- qr.Q(qr(matrix(rnorm(n_features^2), n_features)))
  Q %*% shared + 0.3 * matrix(rnorm(n_features * n_obs), n_features, n_obs)
})

new_result <- apply_alignment(result, AlignmentData(new_sub))
names(get_aligned(new_result))
```

## Key concepts

### Data convention

neuralign uses `features x observations` matrices throughout. Transforms are
operators that **left-multiply** data: `aligned = T %*% original`.

### Reference selection

The *reference* determines what subjects are aligned to. Options include:

- **`"consensus"`** --- iterative mean template (default).
- **`"medoid"`** --- the subject most similar to all others.
- **A subject ID** --- align everyone to a specific subject.
- **A matrix** --- an external template.

### Correspondence regimes

Alignment methods require a *correspondence signal* that defines what should
match across subjects. neuralign supports three regimes:

1. **Shared labels**: all subjects have the same labelled features (e.g., task
   conditions). Use standard Procrustes.
2. **Sparse overlap**: subjects share some but not all features. Use
   `harmonize_feature_blocks()` or graph Procrustes.
3. **No shared labels**: use intrinsic geometry or connectivity fingerprints
   (Gromov--Wasserstein methods).

### Cross-validation and leakage prevention

neuralign provides built-in CV along two orthogonal axes:

- **Subject-axis**: LOSO or k-fold over subjects.
- **Observation-axis**: run-wise or blocked-time folds over timepoints.

Leakage detection is automatic: neuralign warns when you evaluate on training
data and provides `assess_leakage_risk()` for comprehensive auditing.

### Plugin registry

Any R package can register a new alignment method that then works with
`fit_alignment()`, cross-validation, composition, quality assessment, and
serialization. See `vignette("extending")` for details.

```{r available}
available_aligners(details = TRUE)
```

## Choosing a method

| Scenario | Recommended method | Why |
|----------|--------------------|-----|
| Shared task labels, want orthogonal transform | `procrustes` | Standard, fast, well-understood |
| Shared labels with graph/spatial constraints | `procrustes_graph` | Adds geometric regularization |
| Kernel-space alignment | `kprocrustes` | Handles nonlinear structure |
| No shared labels, connectivity only | `gw` | Gromov--Wasserstein on distance matrices |
| No shared labels, mixed features + geometry | `fpgw` | Fused partial Gromov--Wasserstein |

## Package ecosystem

neuralign integrates with several companion packages:

| Package | Role |
|---------|------|
| `manifoldalign` | Computational backends (Procrustes, GW, FPGW solvers) |
| `fmrigds` | fMRI-specific data structures and I/O |
| `dkge` | Discriminant Kernel Generalized Embedding |
| `neuroim2` | Neuroimaging data containers |

## Vignette guide

| Vignette | Topic |
|----------|-------|
| **Getting Started** (`neuralign`) | Hands-on walkthrough: data, fitting, CV, quality, persistence |
| **Choosing Anchors and Guidance** (`anchors`) | Correspondence regimes, ROI projectors, geometry |
| **fMRI Alignment Profile** (`fmri_profile`) | Domain-specific recipes for fMRI workflows |
| **Cross-Validation and Leakage** (`cv_leakage`) | Subject- and observation-axis CV, guard TRs, leakage detection |
| **Extending neuralign** (`extending`) | Writing and registering custom alignment methods |
| **Feature Blocks and Subspaces** (`advanced_blocks`) | Block harmonization, identifiability, composition |
