---
title: "Getting Started with neuralign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with neuralign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

neuralign provides a unified interface for neural alignment methods. It standardizes how alignment models are fit, applied, composed, and persisted---regardless of which algorithm or package implements the actual computation.

This vignette covers:

1. Creating input data with `AlignmentData`
2. Fitting alignments with `fit_alignment()`
3. Cross-validation modes
4. Reference selection strategies
5. Applying models to new data
6. Quality assessment

## Setup

```{r setup}
library(neuralign)
```

## Creating Alignment Data

neuralign expects multi-subject data as a named list of matrices, where each matrix is `(features x observations)`:

```{r create-data}
# Simulate 5 subjects with 100 features and 50 timepoints each
set.seed(42)
n_subjects <- 5
n_features <- 100
n_obs <- 50

# Create a shared signal that all subjects see (with noise)
shared_signal <- matrix(rnorm(n_features * n_obs), n_features, n_obs)

data_list <- lapply(1:n_subjects, function(i) {
  # Each subject has the shared signal plus individual noise
  # Also apply a random rotation to simulate misalignment
  Q <- qr.Q(qr(matrix(rnorm(n_features^2), n_features)))
  Q %*% shared_signal + 0.5 * matrix(rnorm(n_features * n_obs), n_features, n_obs)
})
names(data_list) <- paste0("sub-", sprintf("%02d", 1:n_subjects))

# Wrap in AlignmentData
adat <- AlignmentData(data_list)
adat
```

### Optional Metadata

`AlignmentData` can include additional metadata for methods that need it:

```{r metadata, eval=FALSE}
# For methods requiring task structure (e.g., dkge)
adat_with_design <- AlignmentData(
  data_list,
  design = my_design_matrix,
  space = "MNI152"
)

# For graph-based methods
adat_with_geometry <- AlignmentData(
  data_list,
  geometry = adjacency_matrix
)
```

## Basic Alignment

The main function is `fit_alignment()`. By default, it uses Procrustes alignment with a consensus reference:

```{r basic-fit}
result <- fit_alignment(adat, method = "procrustes")
result
```

### Accessing Results

```{r access-results}
# Get the fitted model (transforms + reference)
model <- get_model(result)

# Get aligned data
aligned <- get_aligned(result)
names(aligned)

# Get a specific subject's transform
Q <- get_transform(model, "sub-01")
dim(Q)  # 100x100 orthogonal matrix

# Get quality metrics computed during fitting
get_quality(result)
```

## Reference Selection

The reference determines what subjects are aligned *to*. neuralign supports several strategies:

### Consensus Reference (default)

Iteratively computes a mean template:

```{r ref-consensus}
result <- fit_alignment(adat, method = "procrustes", reference = "consensus")
```

### Medoid Reference

Selects the subject most similar to all others:

```{r ref-medoid}
# Using correlation distance (default)
result <- fit_alignment(adat, method = "procrustes", reference = "medoid")

# Can also select medoid directly
medoid_subj <- select_reference(adat, method = "medoid", distance = "correlation")
medoid_subj
```

### Specific Subject

Use a named subject as the reference:

```{r ref-subject}
result <- fit_alignment(adat, method = "procrustes", reference = "sub-03")
```

## Cross-Validation

Cross-validation is critical for unbiased evaluation. neuralign ensures **reference selection happens within the training fold only**, preventing leakage.

### Leave-One-Subject-Out (LOSO)

```{r cv-loso}
result_cv <- fit_alignment(adat, method = "procrustes", cv = "loso")

# Check CV info
get_cv_info(result_cv)

# Each subject's transform was fit using the OTHER subjects as training
# Reference was selected from training subjects only
```

### K-Fold Cross-Validation

```{r cv-kfold}
result_kfold <- fit_alignment(
  adat,
  method = "procrustes",
  cv = "kfold",
  cv_folds = 3
)

get_cv_info(result_kfold)
```

## Applying to New Data

Once you have a fitted model, you can apply it to new subjects:

```{r apply-new}
# Create a new subject
new_subject <- list(
  "sub-06" = matrix(rnorm(n_features * n_obs), n_features, n_obs)
)
new_adat <- AlignmentData(new_subject)

# Apply the existing model
new_result <- apply_alignment(result, new_adat)

# The new subject now has an aligned version
names(get_aligned(new_result))
```

neuralign will warn if you apply a model to subjects that were in the training set (potential leakage).

## Quality Assessment

Assess alignment quality with multiple metrics:

```{r quality}
# Compute quality metrics
quality <- alignment_quality(result, metrics = c("correlation", "isc"))

# Mean pairwise correlation after alignment
quality$mean_pairwise_correlation

# Inter-subject correlation
quality$mean_isc

# Formatted summary
print_quality_summary(quality)
```

### Comparing Before/After

```{r quality-improvement}
# Compare aligned vs original
quality_compare <- alignment_quality(
  result,
  original = adat,
  metrics = "correlation"
)

quality_compare$correlation_improvement
```

## Persistence

Save and load alignment models:

```{r persistence, eval=FALSE}
# Save with provenance
save_alignment(result, "my_alignment.rds")

# Load later
loaded_model <- load_alignment("my_alignment.rds")

# Apply to new data
apply_alignment(loaded_model, new_data)

# Export for other tools
export_alignment(result, "alignment_export", format = "json")
```

## Available Methods

Check what methods are registered:

```{r methods}
available_aligners(details = TRUE)
```

Methods from external packages can be auto-loaded when requested (if installed and registered).

## Operator Composition

Chain multiple alignments:

```{r compose, eval=FALSE}
# Example: geometric alignment followed by functional
geo_result <- fit_alignment(geo_data, method = "procrustes")
func_result <- fit_alignment(geo_aligned_data, method = "procrustes")

# Compose: apply geo first, then func
composed <- func_result %*% geo_result

# Check compatibility
check_composition(geo_result, func_result)
```

## Registering Custom Aligners

External packages can register their methods. See `?register_aligner` for details.

```{r register, eval=FALSE}
# Minimal example
register_aligner(
  name = "my_method",
  fit_fn = function(data, reference, train_idx = NULL, ...) {
    # ... compute transforms ...
    list(
      transforms = named_list_of_matrices,
      reference_data = reference_matrix,
      space_from = data@space,
      space_to = data@space
    )
  },
  capabilities = list(
    supports_cv = TRUE,
    transform_type = "orthogonal",
    returns = "operator"
  ),
  package = "mypackage"
)
```

## Next Steps

- See `?fit_alignment` for all parameters
- See `?AlignmentModel` for model structure details
- See `?alignment_quality` for all quality metrics
- For fmrigds integration, see `?as_map_family`
